---
title: "Final report"
author: "Group 6"
date: "2025-10-10"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
#loading libraries
library(tidyverse)
library(dplyr)
library(readr)
library(here)
library(MASS)
library(car)
library(patchwork)
library(ggplot2)
library(stringr)
library(tidyr)
library(knitr)
```

It needs to have: 1) set-up -\> this under the beginning part like the
code and dont show it 2) Research Motivation 3) Research Method 4)
Results (models +plot) 5) Theoretical / Practical Implications 6)
Conclusion

## Research motivation

In the highly competitive field of film making, every decision is
important, especially when significant financial investments are at
stake. Key creative decisions, such as a film's runtime, can influence
on how it is received by audiences. This is further complicated by
genre, as audience expectations for a quirky comedy differ for a biopic.
Audience ratings serve as a direct measure of viewer satisfaction. By
analyzing the relationship between runtime, genre, and these ratings, we
can move from industry intuition to data-backed insights. This knowledge
provides filmmakers and producers with a strategic tool for better
aligning their creative ideas with audience preferences, potentially
lowering financial risk and increasing a film's success.

Based on this, this study will focus on the following research question:

**To what extent does movie runtime influence average audience ratings,
and how does this relationship differ across various film genres?**



## Research method

For this research, data from the IMBD database (basic.csv and rating.csv) were merged, cleaned and filtered to include only feature-length films with at least 250 audience votes and runtimes between 30 and 250 minutes, ensuring statistical reliability. Missing values were removed, and films were categorized into four ordered rating levels. The three most frequent genres were identified, and dummy variables were created to enable comparisons across these genres.

Moreover, boxplots and bar charts were used to illustrate the distributions of runtimes and audience ratings across genres, making patterns and outliers easy to detect. These visualizations were chosen because they offer an overview of how runtime relates to rating both within and between genres.

Finally, using dummy variables for the three most frequent genres allowed for straightforward interpretation of genre specific effects. Together, these methods provide a clear and structured approach to examining the assumed relationship between film runtime, genre, and audience ratings.



## Results

```{r results, echo = FALSE}
movies_clean <- read_csv(here("gen", "data_preparation", "output", "movies_clean.csv"), show_col_types = FALSE))

#Runtime versus Rating Category
p3 <- ggplot(movies_clean, aes(x = rating_category, y = runtime_min, fill = rating_category)) +
  geom_boxplot(outlier.alpha = 0.5) +
  scale_fill_brewer(palette = "Greens") +
  labs(
    title = "Runtime (min) by Rating Category",
    x = "Rating Category",
    y = "Runtime (minutes)"
  ) +
  theme(legend.position = "none")

print(p3)

```
This graph shows the overall link between movie runtime and audience ratings. It reflects an overall tendency in which films with higher ratings, such as "Excellent" and "Good," have longer median runtimes than those with "Average" or "Poor" ratings. When all genres are included, this implies a positive relationship between film duration and perceived quality.

```{r results, echo = FALSE}
movies_clean <- read_csv(here("gen", "data_preparation", "output", "movies_clean.csv"), show_col_types = FALSE))

#Detect all dummy columns automatically (adjust when dataset changes)
str_detect(names(movies_clean), "_dummy$")
dummy_cols <- names(movies_clean)[str_detect(names(movies_clean), "_dummy$")]

#Runtime by Rating Category, Faceted by Dummy Genre 
if (length(dummy_cols) > 0) {
  p6 <- ggplot(long_dummy, aes(x = rating_category, y = runtime_min, fill = rating_category)) +
    geom_boxplot(outlier.alpha = 0.5) +
    facet_wrap(~ genre_dummy) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Runtime by Rating Category (Faceted by Top-3 Genres)",
      x = "Rating Category",
      y = "Runtime (min)"
    ) +
    theme(legend.position = "none")
  
  print(p6)
}

```

The second graph breaks this relationship into the top three genres: comedy, drama, and romance. Within each genre, the trend that longer runtimes lead to higher scores holds true. However, it also shows considerable differences between genres, with dramas consistently lasting longer than comedies at all rating levels.

Comparing the two graphs shows how genre influences the relationship between runtime and rating. The first graph shows the general rule that longer films receive higher ratings. The second graph demonstrates that the concept of a "long" or "short" film is genre-specific. For example, a highly scored comedy is much shorter than a highly rated drama, demonstrating that audience expectations for length vary depending on the sort of film.

## Implications

The analysis demonstrates that movies with longer runtimes tend to have
higher ratings across the three most common genres in the dataset. This
relationship appears to be particularly strong for dramas, while
comedies and romance movies also show a positive but weaker effect.
These findings suggest that audiences have a preference for longer
movies, although the strength of this preference varies by genre.
Considering genre-specific patterns can help film producers optimize
movie runtimes to better match audience expectations. As the dataset
updates daily, the exact results may vary over time.

This research has practical implications for creators of movies.
Creating a longer movie logically demands more work than creating a
shorter movie, and is therefore costly. This research shows that
audiences on average do rate longer movies higher than shorter movies,
which is an argument to make longer movies, especially in the drama genre. Therefore, this research
allows filmmakers to make better informed choices when deciding the
runtime of their movie. Further research could investigate whether there is a causal effect of runtime on audience ratings, to inform the choice of filmmakers even better. 



## Conclusion

In conclusion, this research investigated the relationship between the runtime of movies and their audience ratings. The analysis found that longer runtimes are generally associated with higher audience ratings, especially within the drama genre. Comedies and romance films also showed a similar trend, though the association was weaker. 
While the effect was found in all three examined genres, its strength varied, indicating that runtime does not impact all genres equally. Filmmakers should therefore consider the genre of a movie when making  decisions about a film's length.
